### 2018-04-08 Manuscript/chapter4.tex
Intro and methods draft

\chapter{Gene expression responses to interactive stressors of diet quality and viral infection in \textit{Apis mellifera}}

@@

Pollen is the main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honeybees (\citealt{source}, \citealt{source2}). At the individual level, pollen supplies most of the nutrients necessary for physiological development (\citealt{brodschneider}) and is believed to have considerable impact on longevity (\citealt{longevity}). At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen (\citealt{jelly}, \citealt{jelly2}). Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honeybees. In particular, honeybees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{diverse3}). Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas (\citealt{undeveloped}), and beekeepers rank poor nutrition as one of the main reasons for colony losses (\citealt{bkLoss}). Understanding how undiversified diets affect honeybee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world (\citealt{ag}, \citealt{ag2}).

@@

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}). As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers on honeybee mortality rates.

There are several reasons why, in part of our study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honeybees as they rarely face a total lack of pollen (\citealt{DiPasquale}). Second, in studies that compared honeybee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet (\citealt{DiPasquale}). Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced wtih less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}). As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.

@@

@@

>>>>

\chapter{Gene expression responses to diet quality and viral infection in \textit{Apis mellifera}}

@@

Pollen is the main source of nutrition (including proteins, amino acids, lipids, sterols, starch, vitamins, and minerals) in honeybees (\citealt{source}, \citealt{source2}). At the individual level, pollen supplies most of the nutrients necessary for physiological development (\citealt{brodschneider}) and is believed to have considerable impact on longevity (\citealt{longevity}). At the colony level, pollen enables young workers to produce jelly, which then nourishes larvae, drones, older workers, and the queen (\citealt{jelly}, \citealt{jelly2}). Various environmental changes (including urbanization and monoculture crop production) have significantly altered the nutritional profile available to honeybees. In particular, honeybees are confronted with less diverse selections of pollen, which is of concern because mixed-pollen (polyfloral) diets are generally considered healthier than single-pollen (monofloral) diets (\citealt{diverse}, \citealt{diverse2}, \citealt{alaux}). Indeed, reported colony mortality rates are higher in developed land areas compared to undeveloped land areas (\citealt{undeveloped}), and beekeepers rank poor nutrition as one of the main reasons for colony losses (\citealt{bkLoss}). Understanding how undiversified diets affect honeybee health will be crucial to resolve problems that may arise as agriculture continues to intensify throughout the world (\citealt{ag}, \citealt{ag2}).

@@

Although there is growing interest in how viruses and diet quality affect the health and sustainability of honeybees, as well as a recognition that such factors might operate interactively, there are only a small number of experimental studies thus far directed toward elucidating the interactive effects of these two factors in honeybees (\citealt{intNV}, \citealt{intNV2}, \citealt{intNV3}). We recently used laboratory cages and nucleus hive experiments to investigate how these two factors interact (\citealt{adamInt}). As part of our experiment, we specifically studied the interactive effects of IAPV infection and monofloral diet quality on titers and honeybee mortality rates.

There are several reasons why, in this part of our previous study, we focused only on diet quality (monofloral diets) as opposed to diet diversity (monofloral diets versus polyfloral diets). First, when assessing diet diversity, a sugar diet is often used as a control. However, such an experimental design does not reflect real-world conditions for honeybees as they rarely face a total lack of pollen (\citealt{DiPasquale}). Second, in studies that compared honeybee health using monofloral and polyfloral diets at the same time, if the polyfloral diet and one of the high-quality monofloral diets both exhibited similarly beneficial effects, then it was difficult for the authors to assess if the polyfloral diet was better than most of the monofloral diets because of its diversity or because it contained as a subset the high-quality monofloral diet (\citealt{DiPasquale}). Third, colonies used for pollination in agricultural areas (monoculture) face less diversified pollens (according to Brodschneider, 2010). Pollinating areas are currently undergoing landscape alteration and agriculture intensification, and bees are increasingly faced wtih less diversified diets (monoculture) (\citealt{landscape1}, \citealt{brodschneider}). As a result, there is a need to better understand how monofloral diets affect honeybee health as a step toward mitigating the negative impact of human activity on the honeybee population.

Consequently, in our prior study, for our nutrition factor, we examined two monofloral pollen diets, Cistus (Rockrose) and Castanea (Chestnut). Cistus pollen is generally considered less nutritious than Castanea pollen due to its lower levels of protein, amino acids, antioxidants, calcium, and iron (\citealt{DiPasquale}, \citealt{adamInt}). For our virus factor, one level contained bees that were infected with IAPV and another level contained bees that were not infected with IAPV. This experimental design resulted in four treatment groups that allowed us to assess main effects and interactive effects between diet quality and IAPV infection in honeybees.

Mortality rates of honeybees 72 hour post-inoculation differed (or did not?) among the treatment groups (mixed model ANOVA across all treatment groups, df=@@@, @@@; F=@@@; p<@@@). The effect of virus treatment (mixed model ANOVA, df=@@@, @@@; F=@@@; p<@@@), diet treatment (mixed model ANOVA, df=@@@,@@@; F=@@@; p<@@@), and interaction between the two factors (mixed model ANOVA, df=@@@, @@@; F=@@@, p=@@@) did (or did not?) differ. The virus treatment was significant: For a given diet, honeybees exposed to the virus showed significantly higher mortality rate than honeybees not exposed to the virus (Tukey HSD, p<0.05). Without virus exposure, there was only an intermediate reduction in mortality rate for bees fed Castanea pollen? (Tukey HSD, p>0.05). However, with virus exposure, there was a significant reduction in mortality rate for beeds fed Castanea pollen (Tukey HSD, p<0.05). Overall, we discovered that the higher-quality Castanea diet had the ability to significantly reduce mortality in the presence of IAPV infection compared to the lower-quality Cistus diet (Figure \ref{fig:mortalityAll}).

Comment on (Figure \ref{fig:IAPVAll}) @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/mortalityAll}
  \end{framed}
  \caption{Mortality rates between the four treatments. Each treatment consisted of 15 cages each with 35 honeybees.}
  \label{fig:mortalityAll}
\end{figure}

Comments on (Figure \ref{fig:IAPVAll}) and (Figure \ref{fig:IAPVRNASeq}). @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/IAPVAll}
  \end{framed}
  \caption{IAPV titers between the four treatments. Each treatment contains 15 samples (cages).}
  \label{fig:IAPVAll}
\end{figure}

Following up on these phenotypic findings from our previous study, we now aim to understand the corresponding underlying mechanisms. Transcriptomics is one means to achieve this goal. As it stands, there are only a small number of published experiments examining gene expression patterns related to diet effects (\citealt{alaux2}) and IAPV infection effects (\citealt{galbraith}) in honeybees. As far as we know, there are little to no studies investigating honeybee gene expression patterns specifically related to monofloral diets, and litte to no studies investigating gene expression patterns related to the interaction effects of diet in any broad sense and viral innoculation in any broad sense in honeybees.

In this study, we examine how monofloral diets and viral innoculation influence gene expression patterns in honeybees by focusing on four treatment groups (Cistus diet without IAPV exposure, Castanea diet without IAPV exposure, Cistus diet with IAPV exposure, and Castanea diet with IAPV exposure). We conduct RNA-sequencing analysis on a randomly selected subset of the honeybees we used in our previous study (as is further described in our methods section). We then examine pairwise combinations of treatment groups, the main effect of monofloral diet, the main effect of IAPV exposure, and the interactive effect of the two factors on gene expression patterns.

We also compare the main effect of IAPV exposure in our dataset to that obtained in a previous study conducted by Galbraith and colleagues (\citealt{galbraith}). While our study examined honeybees from polyandrous colonies, the Galbraith study examined honeybees from single-drone colonies. As a consequence, our honeybees will have on average about 75\% genetic variance, and the honeybees from the Galbraith study will have about 25\% genetic variance (\citealt{sisters}). We should therefore expect that the Galbraith study may generate data with lower signal:to:noise ratios than our data due to the lower genetic variation between its replicates. At the same time, our honeybees will be more likely to display the health benefits gained from increased genotypic variation within colonies, including decreased parasitic load (\citealt{multParasite}), increased tolerance to environmental changes (\citealt{divHyp2}), and increasead colony performance (\citealt{geneticDiverse}, \citealt{geneticDiverse2}). Given that honeybees are naturally very polyandrous (\citealt{patriline}), our honeybees may also reflect more realistic environmental and genetic simulations. Taken together, each study provides a different point of value: Our study likely presents less artificial data while the Galbraith data likely presents less messy data. We wish to examine how the gene expression effects of IAPV innoculation compare between these two studies that used such different experimental designs. Namely, we use visualization techniques to assess the signal:to:noise ratios between these two datasets, and differential gene expression (DEG) analyses to assess any overlaps between these two datasets.

@@

The procedures we used to prepare virus inoculum, infect and feed caged honeybees, and quantify IAPV can be reviewed in our previous work (\citealt{adamInt}). The statistical analysis we used to study the main and interaction effects of the two factors on mortality and IAPV titers is also described in our earlier report (\citealt{adamInt}).

\subsection{RNA extraction}

Fifteen cages per treatment were originally sampled. Six live honeybees from each cage were randomly selected 36 hours post inoculation and placed into tubes. Tubes were kept on dry ice and then transferred into a -80C freezer until processing. Eight cages were randomly selected from the original 15 cages, and 2 honeybees per cage were randomly selected from the original six live honeybees per cage. RNA from each pool of two honeybees were extracted using Qiagen RNeasy MiniKit followed by Qiagen DNase treatment. Samples were suspended in water to 200-400 ng/$\mu$l. All samples were then tested on a Bioanalyzer at the DNA core facility to ensure quality (RIN>8).

\subsection{Gene expression}

Samples were sequenced starting on January 14, 2016 at the Iowa State University DNA Facility (Platform: Illumina HiSeq Sequencing; Category: Single End 100 cycle sequencing). Reads were aligned to the BeeBase Version 3.2 genome (\citealt{hbGenome}) from the Hymenoptera Genome Database (\citealt{hymenopteraDB}) using the programs GMAP and GSNAP (\citealt{gsnap}). We tested all six pairwise combinations of treatments for DEGs (pairwise DEGs). We also tested the diet main effect (diet DEGs), virus main effect (virus DEGs), and interaction term for DEGs (interaction DEGs). We then also tested for virus main effect DEGs (virus DEGs) in public data derived from a previous study exploring the gene expression of IAPV virus infection in honeybees (\citealt{galbraith}). We tested each DEG analysis using recommended parameters with DESeq2 (\citealt{deseq2}), edgeR (\citealt{edger}), and LimmaVoom (\citealt{limma}). In all cases, we used a false discovery rate (FDR) threshold of 0.05 (\citealt{benjamini}). Fisher's exact test was used to determine significant overlaps between DEG sets (whether from the same dataset but across different analysis pipelines or from different datasets across the same analysis pipelines).

\subsection{Visualization}

We used @@@ visualization tools from @@@ and visual inference techniques to assess the signal:to:noise ratio in the datasets and to assess the suitability of the DEG calls.

\subsection{Gene Ontology}

DEGs were uploaded as a background list to DAVID Bioinformatics Resources 6.7 (\citealt{davidBio}, \citealt{davidBio2}). The overrepresented gene ontology (GO) terms of DEGs were identified using the BEEBASE\_ID identifier. To fine-tune the GO term list, only significant terms (FDR < 0.05) and those correlating to Biological Processes were considered. The refined GO term list was then imported into REVIGO (\citealt{revigo}), which uses semantic similarity measures to cluster long lists of GO terms.

@@

When statistically analyzing the subset of this data that was used for RNA-sequencing analysis, we found that the mortality rates across diet quality and virus exposure at least numerically retained the same trends (Figure \ref{fig:mortalityRNASeq}).

Comments on (Figure \ref{fig:mortalityRNASeq}) and (Figure \ref{fig:IAPVRNASeq}). @@@

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/mortalityRNASeq}
  \end{framed}
  \caption{Mortality rates between the four treatments. Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages). Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
  \label{fig:mortalityRNASeq}
\end{figure}

\begin{figure}[H]
\centering
  \begin{framed}
  \includegraphics[width=0.8\textwidth]{Images/IAPVRNASeq}
  \end{framed}
  \caption{IAPV titers between the four treatments. Each treatment here only contains the 6 samples (cages) that were used for RNA-sequencing out of its original 15 samples (cages). Two honeybees were randomly selected from each sample (cage) to create a pooled RNA-seqeuncing sample.}
  \label{fig:IAPVRNASeq}
\end{figure}
